\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\title{Stage 3J Physics Note: Unit-Test Fixture Design Before C++ Implementation}
\author{GL-with-AI project}
\date{June 14, 2026}

\begin{document}
\maketitle

\section{Scope and Status}

Stage 3J is a design stage. It does not implement C++ tests, finite
differences, or modified-cartoon CCZ4 source terms. Its purpose is to define
the fixture layer that should exist before project-specific source terms are
implemented.

The target bookkeeping remains
\[
  \mathrm{GR\_SPACEDIM}=4,\qquad
  \mathrm{CH\_SPACEDIM}=2,\qquad
  N_{\rm hidden}=2.
\]
The hidden variable \(h_{ww}\) is a Cartesian-like conformal metric component,
not \(g_{\theta\theta}\). The spherical/cartoon reconstruction supplies the
external factor \(x^2\), and all trace and trace-free projections are full
four-spatial-dimensional operations with denominator \(4\).

\section{Fixture Classes}

The fixture suite should separate exact symbolic checks, C++ algebra unit
tests, RHS-block tests, reference-code comparisons, and convergence or
evolution-level tests. A green result in one class should not be read as
physical correctness of the whole evolution system.

Stage 3J classifies tests as exact, roundoff-level, tolerance-based,
reference-based, or convergence-based. Numeric tolerances are intentionally not
chosen here. Future C++ tolerances must be justified from truncation order,
resolution, floating-point roundoff, interpolation or ghost-zone treatment, and
reference-data uncertainty. They must not be arbitrary thresholds chosen only
because a run passes. Exact symbolic identities should remain exact in the
symbolic scripts.

\subsection{Existing Symbolic Gates}

The current repository already contains symbolic gates for:

\begin{itemize}[leftmargin=2em]
\item Stage 3C flat cartoon geometry, GP extrinsic curvature, and hidden trace
  multiplicity;
\item Stage 3D diagonal nontrivial \(\gamma_{ww}=q(x,z)\) Christoffel and
  Ricci checks;
\item Stage 3E nonconstant \(q(x,z)\) warped-product Ricci derivative-sector
  regression;
\item Stage 3F diagonal conformal-cartoon algebra and determinant/trace
  guards;
\item Stage 3G off-diagonal conformal-cartoon determinant, inverse, trace,
  \(K_{xz}\), and diagonal-limit checks;
\item Stage 3G sheared-flat off-diagonal Ricci gate;
\item Stage 3I small-\(x\) removable-limit and assembled
  \(\hat{\Gamma}^{x}\) axis guard.
\end{itemize}

These symbolic gates should become fixture generators or expected-value
references for the future C++ tests.

\subsection{Future C++ Unit Tests}

The first C++ tests should isolate small blocks:

\begin{itemize}[leftmargin=2em]
\item determinant enforcement with hidden \(h_{ww}\);
\item full 4D trace-free projection with hidden multiplicity;
\item the load-bearing \(1/4\) trace denominator;
\item off-diagonal \(h_{xz}\) determinant and inverse handling;
\item \(K_{ij}\) reconstruction, including
  \[
    K_{xz}=\chi^{-1}\left(A_{xz}+\frac14 h_{xz}K\right);
  \]
\item cartoon Ricci source blocks;
\item a round two-sphere positive-curvature geometry anchor,
  \[
    R_{S^2}=\frac{2}{a^2},
  \]
  or the equivalent Ricci/scalar-curvature statement in the project's angular
  extraction convention;
\item regularized small-\(x\) combinations;
\item assembled \(\tilde{\Gamma}^{x}\) and
  \(\hat{\Gamma}^{x}=\tilde{\Gamma}^{x}+2Z^{x}\);
\item Gamma-driver and \(\hat{\Gamma}^{A}\) ownership boundaries.
\end{itemize}

Each off-diagonal fixture should include a diagonal-limit regression. Each
axis-limit fixture should include both regular Taylor data and deliberately
irregular data that exposes the expected singular behavior.

\section{Pre-C++ Gates}

Before Stage 3K turns this design into an implementation plan, the minimum
gates are:

\begin{itemize}[leftmargin=2em]
\item all Stage 3C--3I symbolic scripts pass;
\item no stale stage-label or old conformal-cartoon filename references remain;
\item determinant, trace, \(h_{ww}\), and \(1/4\) conventions are documented;
\item \(\hat{\Gamma}^{A}=\tilde{\Gamma}^{A}+2Z^{A}\) is documented as the
  connection-sector blueprint;
\item small-\(x\) axis risks, the co-located physical singularity, and
  turduckening/interior regularization are documented as distinct issues;
\item unresolved GRChombo-facing convention choices are carried forward.
\end{itemize}

\section{Reference and Evolution Anchors}

Flat data are necessary but not sufficient. The future validation suite should
also include sheared-flat zero Ricci/source checks, gauge-fixed uniform
GP-string stationarity outside the turduckening region, a carefully scoped
uniform Schwarzschild/black-string anchor, linear Gregory--Laflamme
threshold/growth comparison after convention matching, constraint-violation
damping injection tests, later external/reference implementation comparisons,
and resolution self-convergence.

The Schwarzschild \(\times S^1\) anchor is slicing-dependent. In a GP-type
slice the spatial metric is flat times \(S^1\), while the nontrivial
black-hole geometry is carried by \(K_{ij}\). This gives a spatial-Ricci
geometry check for the GP slice, not a claim that every Schwarzschild spatial
slice has \(R_{ij}=0\). A full spacetime Ricci-flatness check for
Schwarzschild\(_4\times S^1\) is out of scope for the current spatial-cartoon
fixture layer until a spacetime-Ricci symbolic engine exists.

The GL spectrum is a required physical-sector milestone, but it is not a
closed-form elementary unit test. It is a semi-analytic or literature/numerical
spectral anchor whose use requires matching radius convention, periodicity,
perturbation sector, gauge, extraction variable, and slicing/conformal-variable
convention. It validates the physical linearized RHS, including metric,
\(K\), \(A_{ij}\), gauge/source coupling, and cartoon-sector terms.

Constraint-damping validation is a separate required milestone. Since the
physical GL mode is expected to be constraint-satisfying in the continuum, it is
necessary but not sufficient for damping signs. A non-vacuous damping test
should inject controlled linearized Hamiltonian, momentum, \(\Theta\), and
encoded-\(Z^i\) violations around flat or uniform-string data, with
\[
  \hat{\Gamma}^i=\tilde{\Gamma}^i+2Z^i.
\]
Where feasible, the expected linearized CCZ4 constraint-subsystem decay
rates/signs for the chosen \(\kappa_1,\kappa_2\), gauge, and background should
be derived or documented. Otherwise this fixture remains high risk and
reference/formulation-derived.

\section{Unresolved Confirmations}

The following remain open before implementation:

\begin{itemize}[leftmargin=2em]
\item determinant and trace-free enforcement with hidden \(h_{ww}\) in the
  GRChombo-facing project variable set;
\item exact \(\tilde{\Gamma}^{x}\) sign and full \(\hat{\Gamma}^{A}\)
  convention in the GRChombo-facing cartoon extension;
\item Gamma-driver and \(\hat{\Gamma}^{A}\) coupling ownership in the
  GRChombo-facing implementation;
\item gauge choice for uniform-string stationarity;
\item conventions for GL dispersion comparison;
\item which external/reference comparisons are term-by-term and which are only
  trajectory-level validation.
\end{itemize}

\section{Non-Goals}

Stage 3J does not implement C++ tests or source terms, choose final production
tolerances, settle convention questions, or replace the Stage 3K minimal C++
implementation plan.

\end{document}
