\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\title{Stage 3K Physics Note: Minimal C++ Implementation Plan}
\author{GL-with-AI project}
\date{June 14, 2026}

\begin{document}
\maketitle

\section{Scope and Status}

Stage 3K is a planning stage. It does not implement C++, finite differences,
source terms, or production evolutions. Future implementation should live in
repo-owned code, likely under \texttt{code/BlackStringToy/} or a new
repo-owned layer copied into scratch GRChombo trees under ignored
\texttt{runs/}. It must not modify \texttt{external/GRChombo}.

Compilation is not evidence of physical correctness. Algebra-helper fixtures
passing is not integration correctness, and green helper tests do not prove
that variables are read from the correct grid slots. No Stage 4A result should
be described as ``cartoon evolution works'' or ``CCZ4 works.'' Every
implementation stage must have a validation gate.

Stage 4A C++ work requires explicit user approval after Stage 3K review.
Stage 4B and later stages also require explicit approval before source edits.

\section{Convention Policy}

The implementation convention target is GRChombo compatibility plus internal
consistency. External implementations may be used later for validation, but
they are not the source of truth for variable naming, evolved-variable basis,
trace convention, or cartoon-extension bookkeeping. When public GRChombo does
not define the reduced-cartoon extension directly, the project must document
the extension explicitly and use it consistently.

\section{Recommended First Stage}

The recommended first C++ implementation stage is Stage 4A: a repo-owned conformal-cartoon algebra
helper layer plus local, non-grid fixture tests. It must not connect that
helper to grid variables or time evolution. This is the least risky first stage
because it tests the bookkeeping that all later source terms depend on while
leaving the public GRChombo RHS untouched.

Stage 4A should cover:

\begin{itemize}[leftmargin=2em]
\item the off-diagonal determinant
  \[
    (h_{xx}h_{zz}-h_{xz}^{2})h_{ww}^{2};
  \]
\item the reduced inverse conformal metric;
\item full 4D trace and trace-free projection with hidden multiplicity
  \(N_{\rm hidden}=2\);
\item the \(\mathrm{GR\_SPACEDIM}=4\) trace denominator \(1/4\);
\item reconstruction of \(K_{ij}\), including
  \[
    K_{xz}
    =
    \chi^{-1}\left(A_{xz}+\frac14 h_{xz}K\right);
  \]
\item the diagonal limit back to the Stage 3F formulas.
\end{itemize}

Stage 4A explicitly excludes cartoon Ricci helpers, small-axis helpers,
grid-variable wiring, CCZ4 RHS terms, gauge/Gamma-driver terms, constraint
damping, finite-difference stencils, and evolution. Passing Stage 4A only shows
that local algebra formulas were implemented correctly on supplied local
values.

The Stage 4A review follow-up keeps that scope unchanged while making the
local fixture more robust: the conformal-factor power should be expressed as
\(-1/\mathrm{GR\_SPACEDIM}=-1/4\) using the physical spatial dimension, local
floating-point zero guards should use tolerances, and at least one
independently derived \(K_{ij}\) reconstruction oracle should be included.

\section{Stage 4B Public Layout Baseline}

Before editing variable enums or wiring helpers to grid variables, perform a
read-only baseline check of the public CCZ4 variables retained for comparison.
Stage 4B proves that this public GRChombo-facing layout has not drifted
unexpectedly. It does not by itself prove that \(h_{ww}\) or \(A_{ww}\) are
correctly placed.

Stage 4C adds the real hidden enum names and header-level checks in
\texttt{UserVariables.hpp}. With the Stage 4C names \(c_{\rm hww}\) and
\(c_{\rm Aww}\), the assertions are
\[
  c_{\rm hww}=c_K-1,\qquad c_{\rm Aww}=c_\Theta-1,
\]
using compile-time assertions in the variable header. Stage 4D remains the
next narrow step, and only after Stage 4C passes review and gets explicit
approval: give \(h_{ww}\) and \(A_{ww}\) finite scaffold values so the cheap
smoke run no longer dies immediately from NaNs. This is not a physical
evolution check. A broader helper-to-grid handoff remains a later stage.

\section{Deferred Blocks}

Stage 4A should not implement the full CCZ4 RHS, gauge or Gamma-driver
equations, constraint damping, black-string initial data, turduckening,
finite-difference axis treatment, apparent-horizon finding, waveform
extraction, production runs, or comparison with external/reference
implementations.

\section{Validation Map}

\begin{center}
\begin{tabular}{lll}
\toprule
Stage & Required validation & Source stage \\
\midrule
Stage 4A determinant/inverse & normalized determinant, block inverse & 3F--3G \\
Stage 4A trace projection & hidden multiplicity and \(1/4\) guard & 3F--3J \\
Stage 4A \(K_{ij}\) reconstruction & \(K_{xx},K_{xz},K_{zz},K_{ww}\) round trip & 3F--3G \\
Stage 4B public baseline layout & public CCZ4 visible enum/name order & 3J--3K \\
Stage 4C hidden enum placement & header-level \(h_{ww}/A_{ww}\) assertions & 3J--3K \\
Stage 4D finite scaffold values & \(h_{ww}/A_{ww}\) no longer fail cheap smoke with NaNs & 3J--3K \\
Later grid handoff & helper input slots receive intended live components & 3J--3K \\
Stage 4E cartoon Ricci helper & flat, round \(S^2\), warped, sheared-flat gates & 3C--3G, 3J \\
Stage 4F small-axis helper & assembled \(\hat{\Gamma}^x\) limit guard & 3I--3J \\
Later constraint damping & injected-constraint linearized milestone & 3H--3J \\
\bottomrule
\end{tabular}
\end{center}

\section{Primary Risks}

The first implementation must guard against using a 2D trace instead of a full
4D trace, omitting \(h_{ww}\) from determinant or trace operations,
double-counting the external \(x^2\) factor by treating \(h_{ww}\) as
\(g_{\theta\theta}\), using reciprocal inverses when \(h_{xz}\ne0\), leaving a
residual \(1/x\) in \(\hat{\Gamma}^{x}\), trusting flat-zero checks for
constraint damping, wiring correct helper formulas to the wrong enum slot,
editing \texttt{external/GRChombo}, or trusting compilation alone.

\section{Acceptance}

Stage 3K is complete when the minimal first implementation stage is limited to local
algebra helpers and non-grid fixtures, deferred blocks are listed, repo-owned
implementation targets are mapped without editing code, each stage is linked to
Stage 3J tests, automated layout gating is required before grid wiring, and no
C++ implementation has started. Stage 4A is the approved follow-up stage that
implements the local helper and non-grid fixture only.

\end{document}
