\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}

\title{Stage 3G Physics Note: Off-Diagonal Conformal Cartoon Translation}
\author{GL-with-AI Project}
\date{\today}

\begin{document}
\maketitle

\section{Scope}

Stage 3G extends the Stage 3F diagonal conformal-cartoon translation to a
reduced metric block with nonzero \(h_{xz}\) or \(\gamma_{xz}\). It is a
derivation and symbolic-validation note only. No C++ source terms or variables
are implemented, and symbolic execution alone does not establish physical
correctness of an evolution.

\section{Metric Ansatz}

The reduced physical spatial metric is
\begin{equation}
  dl^2
  =
  \gamma_{xx}\,dx^2
  +
  2\gamma_{xz}\,dx\,dz
  +
  \gamma_{zz}\,dz^2
  +
  x^2\gamma_{ww}\,d\Omega_2^2 .
\end{equation}
The gridded directions are \(x,z\). The hidden transverse directions are two
equivalent \(w\) directions, so
\begin{equation}
  N_{\rm hidden}
  =
  \mathrm{GR\_SPACEDIM}-\mathrm{CH\_SPACEDIM}
  =
  4-2=2.
\end{equation}
As before, \(h_{ww}\) and \(\gamma_{ww}\) are hidden Cartesian-like components;
the spherical/cartoon reconstruction supplies the external factor \(x^2\).

\section{Determinants}

The reduced physical block determinant is
\begin{equation}
  D_\gamma
  =
  \gamma_{xx}\gamma_{zz}-\gamma_{xz}^2.
\end{equation}
Thus the cartoon/Cartesian-like four-dimensional physical determinant is
\begin{equation}
  \det\gamma_{\rm 4D}
  =
  D_\gamma\,\gamma_{ww}^2.
\end{equation}
With
\begin{equation}
  h_{ij}=\chi\gamma_{ij},
  \qquad
  \gamma_{ij}=\chi^{-1}h_{ij},
\end{equation}
the conformal factor is
\begin{equation}
  \chi
  =
  \left[
    \left(\gamma_{xx}\gamma_{zz}-\gamma_{xz}^2\right)
    \gamma_{ww}^2
  \right]^{-1/4}.
\end{equation}
For the conformal block,
\begin{equation}
  D_h
  =
  h_{xx}h_{zz}-h_{xz}^2,
\end{equation}
and the conformal determinant condition is
\begin{equation}
  \det h_{\rm 4D}
  =
  D_h h_{ww}^2
  =
  \left(h_{xx}h_{zz}-h_{xz}^2\right)h_{ww}^2
  =
  1.
\end{equation}
This determinant is imposed on the reduced cartoon/Cartesian-like variables;
it does not include spherical-coordinate Jacobian factors.

\section{Inverse Metric}

The inverse conformal block is
\begin{equation}
  h^{xx}=\frac{h_{zz}}{D_h},
  \qquad
  h^{xz}=-\frac{h_{xz}}{D_h},
  \qquad
  h^{zz}=\frac{h_{xx}}{D_h},
  \qquad
  h^{ww}=\frac{1}{h_{ww}}.
\end{equation}

\section{Tracelessness}

The conformal tracelessness condition is the full four-dimensional one,
\begin{equation}
  h^{xx}A_{xx}
  +
  2h^{xz}A_{xz}
  +
  h^{zz}A_{zz}
  +
  2h^{ww}A_{ww}
  =
  0.
\end{equation}
The \(2h^{xz}A_{xz}\) term is the symmetric off-diagonal contraction. The
\(2h^{ww}A_{ww}\) term is the hidden-direction multiplicity. Solving for the
hidden component gives
\begin{equation}
  A_{ww}
  =
  -\frac{h_{ww}}{2}
  \left(
    h^{xx}A_{xx}
    +
    2h^{xz}A_{xz}
    +
    h^{zz}A_{zz}
  \right).
\end{equation}

\section{Extrinsic Curvature}

For \(\mathrm{GR\_SPACEDIM}=4\),
\begin{equation}
  K_{ij}
  =
  \chi^{-1}
  \left(A_{ij}+\frac{h_{ij}K}{4}\right).
\end{equation}
The off-diagonal component is therefore
\begin{equation}
  K_{xz}
  =
  \chi^{-1}
  \left(A_{xz}+\frac{h_{xz}K}{4}\right).
\end{equation}
The trace is
\begin{equation}
  K
  =
  \gamma^{xx}K_{xx}
  +
  2\gamma^{xz}K_{xz}
  +
  \gamma^{zz}K_{zz}
  +
  2\gamma^{ww}K_{ww}.
\end{equation}
The denominator \(4\) is the full physical spatial dimension, not the two
gridded directions and not ordinary \(3+1\) intuition.

\section{Script Checks}

The repository script
\texttt{docs/derivations/offdiagonal\_conformal\_cartoon\_sympy.py}
checks the physical and conformal determinant identities, the inverse metric,
the explicit normalized \(\det h_{\rm 4D}=1\) guard, full four-dimensional
tracelessness, reconstruction of
\(K_{xx},K_{xz},K_{zz},K_{ww}\), and a denominator guard showing that \(p=4\)
works while \(p=2\) and \(p=3\) fail under a numeric substitution.

It also checks the diagonal limit
\begin{equation}
  h_{xz}=\gamma_{xz}=A_{xz}=K_{xz}=0,
\end{equation}
which recovers the Stage 3F diagonal formulas.

\section{Companion Ricci Gate}

The companion repository script
\texttt{docs/derivations/offdiagonal\_ricci\_flat\_gate\_sympy.py}
checks the geometric Ricci engine on flat metrics with genuine off-diagonal
cross terms. The constant-shear case is
\begin{equation}
  dl^2
  =
  (1+\lambda^2)\,dx^2
  +
  2\lambda\,dx\,dz
  +
  dz^2
  +
  x^2d\Omega_2^2,
\end{equation}
corresponding to \(dZ=dz+\lambda dx\). The \(x\)-dependent shear case is
\begin{equation}
  dl^2
  =
  (1+x^2)\,dx^2
  +
  2x\,dx\,dz
  +
  dz^2
  +
  x^2d\Omega_2^2,
\end{equation}
corresponding to \(dZ=dz+x\,dx\). The script verifies that the full matrix
inverse is used and that all Ricci tensor components and the Ricci scalar
vanish for both metrics. This is a targeted off-diagonal Ricci regression, not
a full CCZ4 RHS validation.

\section{Limitations}

This note is an algebraic translation and validation scaffold. It does not
derive the full CCZ4 right-hand sides, does not address regularity near
\(x=0\), and does not add or validate any C++ implementation.

\end{document}
