\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}

\title{Stage 3E Physics Note: Nonconstant Hidden-Metric Ricci Regression}
\author{GL-with-AI Project}
\date{\today}

\begin{document}
\maketitle

\section{Scope}

This note summarizes the Stage 3E symbolic regression for nonconstant
\(\gamma_{ww}=q(x,z)\). It is a derivation artifact only. It does not implement
CCZ4 source terms or add GRChombo variables.

\section{Metric Ansatz}

The regression uses
\begin{equation}
  dl^2 = dx^2 + dz^2 + f(x,z)^2 d\Omega_2^2,
\end{equation}
with
\begin{equation}
  f=xP,\qquad
  P=1+\lambda x+\mu z,\qquad
  q=P^2.
\end{equation}
Equivalently,
\begin{equation}
  dl^2 = dx^2 + dz^2 + x^2q(x,z)d\Omega_2^2.
\end{equation}
Here \(q=\gamma_{ww}\) is the hidden cartoon component; the factor \(x^2\) is a
coordinate reconstruction factor.

\section{Warped-product Reference}

For a two-dimensional unit-sphere fiber over the flat base \((x,z)\), the
independent reference formulas are
\begin{equation}
  R_{AB}=-2f^{-1}\partial_A\partial_B f,
  \qquad A,B\in\{x,z\},
\end{equation}
and
\begin{align}
  R_{\theta\theta}
  &=
  1-|\nabla f|^2-f\Delta f,\\
  R_{\phi\phi}
  &=
  \left(1-|\nabla f|^2-f\Delta f\right)\sin^2\theta,
\end{align}
where
\begin{equation}
  |\nabla f|^2=f_x^2+f_z^2,
  \qquad
  \Delta f=f_{xx}+f_{zz}.
\end{equation}
The Ricci scalar reference is
\begin{equation}
  R=-4\frac{\Delta f}{f}
  +2\frac{1-|\nabla f|^2}{f^2}.
\end{equation}

\section{Regression Checks}

The SymPy script computes Ricci directly from the metric and asserts agreement
with the warped-product reference for \(R_{xx}\), \(R_{zz}\), \(R_{xz}\),
\(R_{\theta\theta}\), \(R_{\phi\phi}\), and \(R\). It also asserts that
\(\partial_x q\), \(\partial_z q\), and \(\partial_x\partial_z q\) are not
identically zero, verifies the flat special case \(\lambda=\mu=0\), and checks
the nontrivial substitution \(\lambda=2,\mu=3\).

\section{Caveats}

The Stage 3E script recomputes Ricci using its own copy of symbolic engine
helpers such as \texttt{christoffel\_symbols} and \texttt{ricci\_tensor}. It
therefore directly guards the Ricci engine on a nonconstant warped-product case
with active \(q\)-derivative terms. It does not directly substitute the
nonconstant profile into the exact abstract Ricci expressions printed by the
Stage 3D script. The Stage 3D expressions are validated transitively because
they were produced by the same kind of engine, which is now checked on flat
limits, the constant-\(q_0\) cone, and this nonconstant warped-product
regression. The residual risk is that a wiring mistake could corrupt the
printed or assembled Stage 3D abstract expression while leaving the engine
functions intact.

The derivation scripts now duplicate symbolic engine helper functions. This is
acceptable short-term, but a future cleanup should factor common helpers into a
shared module such as \texttt{docs/derivations/cartoon\_geometry\_common.py}.

For the chosen profile \(f=x(1+\lambda x+\mu z)\), one has \(f_{zz}=0\), so the
warped-product formula gives
\begin{equation}
  R_{zz}=-2f^{-1}f_{zz}=0.
\end{equation}
This does not make the \(z\)-derivative sector vacuous: \(\partial_z q\) is
nonzero, mixed derivative structure is active, \(R_{xz}=-2\mu/f\) is generally
nonzero, and the angular Ricci components contain \(\mu^2\) terms through
\(|\nabla f|^2\). The narrower limitation is that a bug appearing only in
\(R_{zz}\) when \(f_{zz}\ne0\) would not be caught by this profile. A future
optional refinement could use a profile quadratic in \(z\), making \(R_{zz}\)
a nonzero target.

\section{Limitations}

This is still not a full CCZ4 RHS derivation. Off-diagonal \(g_{xz}\),
non-flat base metric components, direct substitution into the exact Stage 3D
printed abstract Ricci expression, a nonconstant profile with \(f_{zz}\ne0\),
conformal variables \(\chi,h_{ww}\), regularized small-\(x\) limits, C++
translation, unit-test fixtures, and a shared symbolic-geometry helper module
remain future work.

\end{document}
